\chapter{Evaluation}
\label{chap:evaluation}

\TODO{Repeat the Methodology sections, this time doing the actual comparison.}

\section{Portability}

\TODO{High-level portability score}

\subsection{Platform support}

\paragraph{Android \& iOS} \TODO{no additional platforms}

\paragraph{Apache Cordova} \TODO{For now: Windows Phone, BlackBerry. For later: Tizen, Qt, Firefox Mobile OS, Ubuntu Mobile, Windows (Desktop)}

\paragraph{Motorola Rhodes} \TODO{Windows Phone, BlackBerry}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Toolset reuse}

\paragraph{Android \& iOS} \TODO{Needed: Mac computer, Xcode, Eclipse}

\paragraph{Apache Cordova} \TODO{Any operating system with a text editor or web IDE, cloud builder available}

\paragraph{Motorola Rhodes} \TODO{Any operating system with a Ruby IDE, cloud builder available}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Code reuse}

\paragraph{Android \& iOS} \TODO{Absolutely nothing!}

\paragraph{Apache Cordova} \TODO{100\%, Except for native plugins: 0\%}

\paragraph{Motorola Rhodes} \TODO{Any operating system with a Ruby IDE, cloud builder available}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Portability Summary?}

\TODO{summary with all portability values?}

\section{Application Experience}

\TODO{High-level Application Experience score}

\subsection{Native Integration}

\paragraph{Android \& iOS} \TODO{list native device APIs}

\paragraph{Apache Cordova} \TODO{list native device APIs}

\paragraph{Motorola Rhodes} \TODO{list native device APIs}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{UI Capabilities}

\paragraph{Android \& iOS} \TODO{list UI elements \& styling options}

\paragraph{Apache Cordova} \TODO{list default UI (form) elements in HTML5}

\paragraph{Motorola Rhodes} \TODO{same as Cordova minus HTML5 polyfills!}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Performance}

\TODO{subjective testing allowed? AHP can make up for it?}

\paragraph{Android \& iOS} \TODO{Best}

\paragraph{Apache Cordova} \TODO{average/worse than average}

\paragraph{Motorola Rhodes} \TODO{Worst}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Application Experience Summary?}

\section{Productivity}

\TODO{High-level productivity score}

\subsection{Skill reuse}

\paragraph{Android \& iOS} \TODO{Android is Java, iOS is Objective C. Java is known from EE, objective C is not. about 50\%}

\paragraph{Apache Cordova} \TODO{Cordova is HTML + JS + CSS, completely known: 100\%}

\paragraph{Motorola Rhodes} \TODO{Ruby + HTML + JS + CSS, not completely known: 40--50\%}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Tooling}

\paragraph{Android \& iOS} \TODO{Great IDE for iOS, Good Eclipse plugin for Android}

\paragraph{Apache Cordova} \TODO{Few or no tools (Ripple emulator maybe?), Cloud builder?}

\paragraph{Motorola Rhodes} \TODO{Awful (out of the box), cloud builder?}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Testing}

\paragraph{Android \& iOS} \TODO{}

\paragraph{Apache Cordova} \TODO{}

\paragraph{Motorola Rhodes} \TODO{}

\paragraph{Verdict} \TODO{pairwise scores in table}

\subsection{Productivity Summary?}

\section{Global verdict}
